# Screen time: effects on sleep and mood

Author: Niklas Edvall <br> Date: 2026-01-14

## Introduction

This report presents a mock-up cross-sectional survey study designed for instructional purposes. The dataset consists of 120 simulated respondents spanning a wide age range and representing students and staff within an academic setting.

The survey includes basic demographic information, lifestyle factors such as daily screen time and sleep duration, and two brief, widely used mental health questionnaires: the [WHO-5 Well-Being Index](https://www.who.int/publications/m/item/WHO-UCN-MSD-MHE-2024.01) and the [PHQ-4](https://pubmed.ncbi.nlm.nih.gov/19996233/), measuring subjective well-being and symptoms of anxiety and depression, respectively.

The data were simulated to reflect plausible real-world relationships with the primary aim to illustrate common analysis methods, including descriptive analysis, scale construction, correlation, and regression modeling, rather than to draw substantive conclusions.

```{r}
#| echo: false

# Load relevant packages
library(arsenal) # Create descriptive tables
library(ggplot2) # Make nice figures

# Read in data
dat <- read.csv("data/simdata1.csv")

# Recode variables
dat$sex <- factor(dat$sex, levels = c(1,2,3), labels = c("M", "F", "O"))
dat$survey <- factor(dat$survey, levels = c(1,2,3,4), labels = c("Webpage", "Newsletter", "Colleagues", "Lecture"))
dat$position <- factor(dat$position, levels = c(1,2,3,4,5), labels = c("Student", 
                                                                       "PhD Candidate", 
                                                                       "Staff", 
                                                                       "Teacher", 
                                                                       "Other staff"))
```

## Data collection

The survey was advertised to students and staff at a Swedish university and required university login credentials to access. Participation was voluntary, and no financial or other incentives were offered. Respondents completed the questionnaire in a single session, which typically required less than 10 minutes. No personally identifiable information was stored. The dataset used in this report is simulated and intended solely for educational purposes.

## Demographics

Table 1 show age distribution

```{r}
#| results: asis
#| echo: false
# Descriptives
tab1 <- tableby(data = dat, position ~ age + survey, 
                digits = 1, 
                test = FALSE,
                numeric.stats = c("meansd"),
                cat.stats = c("countpct"))
cat('<div style="font-size: 12px;"><style>table td, table th { padding: 2px 8px; line-height: 1.2; }</style>')
summary(tab1, labelTranslations = c(age = "Age (years)", survey = "Survey"))
cat('</div>')
```

## Results

Figures goes here