---
format: 
  html:
    page-layout: full
    max-width: 1400px
---

# Screen time: effects on sleep and mood

Author: Niklas Edvall <br> Date: 2026-01-14

## Introduction

This report presents a mock-up cross-sectional survey study designed for instructional purposes. The dataset consists of 120 simulated respondents spanning a wide age range and representing students and staff within an academic setting.

The survey includes basic demographic information, lifestyle factors such as daily screen time and sleep duration, and two brief, widely used mental health questionnaires: the [WHO-5 Well-Being Index](https://www.who.int/publications/m/item/WHO-UCN-MSD-MHE-2024.01) and the [PHQ-4](https://pubmed.ncbi.nlm.nih.gov/19996233/), measuring subjective well-being and symptoms of anxiety and depression, respectively.

The data were simulated to reflect plausible real-world relationships with the primary aim to illustrate common analysis methods, including descriptive analysis, scale construction, correlation, and regression modeling, rather than to draw substantive conclusions.

```{r}
#| echo: false

# Load relevant packages
library(arsenal) # Create descriptive tables
library(ggplot2) # Make nice figures

# Define text sizes for consistent plotting
plot_title_size <- 18
axis_title_size <- 16
axis_text_size <- 14

# Read in data
dat <- read.csv("data/simdata1.csv")

# Recode variables
dat$sex <- factor(dat$sex, levels = c(1,2,3), labels = c("M", "F", "O"))
dat$survey <- factor(dat$survey, levels = c(1,2,3,4), labels = c("Webpage", "Newsletter", "Colleagues", "Lecture"))
dat$position <- factor(dat$position, levels = c(1,2,3,4,5), labels = c("Student", 
                                                                       "PhD Cand.", 
                                                                       "Researcher", 
                                                                       "Teacher", 
                                                                       "Other staff"))
dat$sleep_cat <- cut(dat$sleep, breaks = c(0,6,9,24), 
                        labels = c("Short (<6)", "Normal (6-9)", "Long (>9)"), right = FALSE)
```

## Data collection

The survey was advertised to students and staff at a Swedish university and required university login credentials to access. Participation was voluntary, and no financial or other incentives were offered. Respondents completed the questionnaire in a single session, which typically required less than 10 minutes. No personally identifiable information was stored. The dataset used in this report is simulated and intended solely for educational purposes.

## Demographics

Table 1 presents the demographic characteristics of the study sample stratified by position within the academic institution. The sample comprised 120 participants with a mean age of 38.4 years (SD = 14.6), ranging from 16 to 74 years. The distribution varied considerably across positions, with students being notably younger (M = 24.3 years) compared to other staff categories. 

Survey recruitment methods showed differential effectiveness, with the majority of participants (59%) recruited through the newsletter, followed by lecture announcements (23%) and colleague referrals (16%). A small proportion of missing data (2.5%) was observed for the survey recruitment variable.

```{r}
#| results: asis
#| echo: false
# Descriptives
tab1 <- tableby(data = dat, position ~ age + survey, 
                digits = 1, 
                test = FALSE,
                numeric.stats = c("meansd", "N"),
                cat.stats = c("countpct", "N"))
cat('<div style="font-size: 16px;"><style>table td, table th { padding: 2px 8px; line-height: 1.2; }</style>')
summary(tab1, labelTranslations = c(age = "Age (years)", survey = "Survey"))
cat('</div>')
```

## Descriptives

Figure 1 illustrates the distribution of daily screen time across different academic positions, stratified by sex. Overall, screen time varied substantially within each position category, with students showing the highest median screen time (around 4.5 hours/day), followed by PhD candidates and researchers. Notable sex differences emerged across most positions, with males generally reporting higher screen time than females.

The second panel examines screen time in relation to sleep duration categories, revealing an inverse relationship whereby individuals with shorter sleep duration (<6 hours) tended to report higher screen time, while those with longer sleep (>9 hours) showed lower screen time patterns.

```{r}
#| echo: FALSE
#| message: FALSE
#| fig-align: center
#| layout-ncol: 2

ggplot(dat[dat$sex != "O", ], aes(x = position, y = screen_time, fill = sex)) +
  geom_boxplot() +
  scale_fill_brewer(type = "qual", palette = "Set2", labels = c("Female", "Male")) +
  labs(
    title = "Screen Time by Position",
    x = "Position",
    y = "Screen Time (hours)",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, size = axis_text_size),
    plot.title = element_text(size = plot_title_size, hjust = 0.5),
    axis.title = element_text(size = axis_title_size),
    axis.title.y = element_text(vjust = 3),
    axis.text = element_text(size = axis_text_size),
    axis.line = element_line(color = "black")
  )

ggplot(dat[dat$sex != "O", ], aes(x = sleep_cat, y = screen_time, fill = sex)) +
  geom_boxplot() +
  scale_fill_brewer(type = "qual", palette = "Set2", labels = c("Female", "Male")) +
  labs(
    title = "Screen Time by Sleep",
    x = "Sleep Category",
    y = "Screen Time (hours)",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, size = axis_text_size),
    plot.title = element_text(size = plot_title_size, hjust = 0.5),
    axis.title = element_text(size = axis_title_size),
    axis.title.y = element_text(vjust = 3),
    axis.text = element_text(size = axis_text_size),
    axis.line = element_line(color = "black")
  )

```

## Results

Figure 2 presents the relationships between age and mental health outcomes. The left panel shows a modest positive association between age and WHO-5 well-being scores (r = `r round(cor(dat$age, dat$WHO5_total), 3)`), suggesting that older participants reported slightly higher subjective well-being. 

Conversely, the right panel reveals a weak negative relationship between age and PHQ-4 scores (r = `r round(cor(dat$age, dat$PHQ4_total), 3)`), indicating a trend toward lower anxiety and depression symptoms with increasing age, though this association did not reach conventional statistical significance. These patterns align with existing literature suggesting potential age-related differences in mental health outcomes within academic settings.

```{r}
#| echo: FALSE
#| message: FALSE
#| fig-align: center
#| layout-ncol: 2

# WHO-5 plot
ggplot(dat, aes(x = WHO5_total, y = age)) +
  geom_point(color = "dark grey") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "WHO-5 Well-being vs Age",
    x = "WHO-5 Well-being Score",
    y = "Age (years)"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black"),
    plot.title = element_text(size = plot_title_size, hjust = 0.5),
    axis.title = element_text(size = axis_title_size),
    axis.title.y = element_text(vjust = 3),
    axis.text = element_text(size = axis_text_size)
  )

# PHQ-4 plot
ggplot(dat, aes(x = PHQ4_total, y = age)) +
  geom_point(color = "dark grey") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "PHQ-4 vs Age",
    x = "PHQ-4 Score",
    y = "Age (years)"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(color = "black"),
    plot.title = element_text(size = plot_title_size, hjust = 0.5),
    axis.title = element_text(size = axis_title_size),
    axis.title.y = element_text(vjust = 3),
    axis.text = element_text(size = axis_text_size)
  )
```

```{r}
#| echo: false

# Fit linear models
model_who5 <- lm(WHO5_total ~ age, data = dat)
model_phq4 <- lm(PHQ4_total ~ age, data = dat)

# Extract model summaries
who5_summary <- summary(model_who5)
phq4_summary <- summary(model_phq4)
```

Linear regression analyses were conducted to quantify the relationships observed in Figure 2. For the WHO-5 well-being model, well-being scores were significantly predicted by age (β = `r round(coef(model_who5)[2], 3)`, SE = `r round(who5_summary$coefficients[2,2], 3)`, t = `r round(who5_summary$coefficients[2,3], 2)`, p = `r round(who5_summary$coefficients[2,4], 3)`, R² = `r round(who5_summary$r.squared, 3)`). 

The PHQ-4 anxiety/depression model showed a marginally significant negative relationship with age (β = `r round(coef(model_phq4)[2], 3)`, SE = `r round(phq4_summary$coefficients[2,2], 3)`, t = `r round(phq4_summary$coefficients[2,3], 2)`, p = `r round(phq4_summary$coefficients[2,4], 3)`, R² = `r round(phq4_summary$r.squared, 3)`). 

These models explain a modest proportion of variance in mental health outcomes, suggesting that while statistically detectable, the practical significance of these age-related differences may be limited.

## Conclusion
This mock-up study illustrates the application of various statistical techniques to explore the relationships between screen time, sleep, and mental health outcomes in an academic population. The findings suggest potential age-related differences in well-being and mental health symptoms, although the effect sizes are modest.